A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2024-11-09 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: typos

In the case of ensembles, typical verifiable statements will be ``the average energy of the particle is between 3 and 4 $eV$,'' ``the probability of getting heads is between 49 and 51 percent,'' ``the probability distribution for the position is a Gaussian of average 0 $m$ and 1 $m$ standard deviation with 1\%.'' Note that these are not results of single shot measurements, and therefore verifiable statements are distinctly different from quantum measurements. In the same manner, these are not measurements that extract one bit of information, as there is no requirement of termination in the negative case.

@@

\subsection{Statistical mixtures}
Another basic requirement is the ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, then $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ is the ensemble that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}. Moreover, since the mixing operation must be consistent with experimental verifiability, it will be topologically continuous an the ensemble space will be a topological convex space.

Ultimately, the convex structure is responsible for all linear structures we have in physics. Both in classical and quantum mechanics, the ensemble space is not just a convex set, but a convex subset of a vector space. In those cases, in fact, the convex structure is, in a sense, invertible. If we fix $\ens$, $\ens[a]$ and $p$ then, if it exists, there is only one $\ens[b]$ such that $\ens = p \ens[a] + (1-p) \ens[b]$. The ensemble space is \textbf{complemented}, meaning that there is only one complement to a component of an ensemble.

This perspective allows to understand what negative probability is. Suppose we have a real vector space of dimensions $n$, then $n$ linearly independent vectors are enough to express every point as a linear combination. If we add another point, we can now express every point as an affine combination, that is a linear combination where the coefficients sum to one, though some can still be negative. Since the space of mixed space in quantum mechanics is a real vector space, we can choose a linearly dependent set of states space span the whole space, and every mixed state can be expressed as an affine combination, as a pseudo-probability.

@@

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined \textbf{entropy}. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following
\begin{enumerate}
	\item strictly convex: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item upper bounded: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.\footnote{It can be shown that if an upper bound that only depends on the mixing coefficients exists, then it must be the Shannon entropy.}

@@

Another important feature is that if two ensembles saturate the upper bound, then they are orthogonal in both classical and quantum mechanics. Therefore we define two ensembles to be \textbf{orthogonal} if their mixture lead to a maximal entropy increase. An additional axiom is valid for orthogonal ensembles. We say an ensemble is a component of another ensemble if the second can be expressed as a mixture of the first with some other ensemble. On physical grounds, two orthogonal ensembles $\ens_1, \ens_2 \in \Ens$ are perfectly distinguishable and must have no common component. That is, there is no $\ens_3 \in \Ens$ that is a component of both $\ens_1$ and $\ens_2$. Note that the converse is not generally true: two pure states in quantum mechanics have no components to have in common, but are not necessarily orthogonal. We find that an ensemble space where no common component implies orthogonality can be described by a space of classical probability measures. Non-classicality, then, is exactly the failure of this property.

>>>>

In the case of ensembles, typical verifiable statements will be ``the average energy of the particle is between 3 and 4 $eV$,'' ``the probability of getting heads is between 49 and 51 percent,'' ``the probability distribution for the position is a Gaussian of average 0 $m$ and 1 $m$ standard deviation within 1\%.'' Note that these are not results of single shot measurements, and therefore verifiable statements are distinctly different from observables in quantum mechanics. In the same manner, these are not measurements that extract one bit of information, as there is no requirement of termination in the negative case.

@@

\subsection{Statistical mixtures}
Another basic requirement is the ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, then $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ is the ensemble that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}. Moreover, since the mixing operation must be consistent with experimental verifiability, it will be topologically continuous and the ensemble space will be a topological convex space.

Ultimately, the convex structure is responsible for all linear structures we have in physics. Both in classical and quantum mechanics, the ensemble space is not just a convex set, but a convex subset of a vector space. In those cases, in fact, the convex structure is, in a sense, invertible. If we fix $\ens$, $\ens[a]$ and $p$ then, if it exists, there is only one $\ens[b]$ such that $\ens = p \ens[a] + (1-p) \ens[b]$. The ensemble space is \textbf{complemented}, meaning that there is only one complement to a component of an ensemble.\footnote{In the context of convex spaces, this property is called ``cancellative.''}

This perspective allows us to understand what negative probability is. Suppose we have a real vector space of dimensions $n$, then $n$ linearly independent vectors are enough to express every point as a linear combination. If we add another point, we can now express every point as an affine combination, that is a linear combination where the coefficients sum to one, though some can still be negative. Since the space of mixed states in quantum mechanics is a real vector space, we can choose a linearly dependent set of states that span the whole space, and every mixed state can be expressed as an affine combination, as a pseudo-probability.

@@

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined \textbf{entropy}. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following
\begin{enumerate}
	\item strictly concave: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item bounded from above: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.\footnote{It can be shown that if an upper bound that only depends on the mixing coefficients exists, then it must be the Shannon entropy.}

@@

Another important feature is that if two ensembles saturate the upper bound, then they are orthogonal in both classical and quantum mechanics. Therefore we define two ensembles to be \textbf{orthogonal} if their mixture leads to a maximal entropy increase. An additional axiom is valid for orthogonal ensembles. We say an ensemble is a component of another ensemble if the second can be expressed as a mixture of the first with some other ensemble. On physical grounds, two orthogonal ensembles $\ens_1, \ens_2 \in \Ens$ are perfectly distinguishable and must have no common component. That is, there is no $\ens_3 \in \Ens$ that is a component of both $\ens_1$ and $\ens_2$. Note that the converse is not generally true: two pure states in quantum mechanics have no components in common, but are not necessarily orthogonal. We find that an ensemble space where no common component implies orthogonality can be described by a space of classical probability measures. Non-classicality, then, is exactly the failure of this property.
